### 2017-07-19 pkg/model4you/inst/JORS/man_model4you.Rnw
first draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1507 255f4249-2263-4417-b1ea-8498779e27a7

<<setup, echo=FALSE>>=
library("knitr")
knitr::opts_chunk$set(cache = FALSE, message = FALSE, fig.align = "center")

@@

\section*{Abstract}
Typical models estimating treatment effects assume that the treatment effect is
the same for all individuals.  Model-based recursive partitioning allows to
relax this assumption and allows for the estimation of stratified treatment
effects (model-based trees) or even personalised treatment effects (model-based
forests).  With model-based trees one can compute treatment effects for
different strata of individuals. The strata are found in a data driven fashion
and depend on characteristics of the individuals.  Model-based random forests
allow for a similarity estimation between individuals in terms of model
estimates. The similarity measure can then be used to estimate personalised
models.  The R package \emph{model4you} implements these stratified and
personalised models with a focus on ease of use and interpretability so that
clinicians and other users can take the model they usually use for the
estimation of the average treatment effect and with a few lines of code get a
visualisation that is easy to understand and interpret.

@@

\paragraph{(1) Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristicum) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data is partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case in clinical trial
in \citep{seibold_model-based_2016}.  Just a single line of code lets the user
compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.  To
compute personalised models for all observations in the training data is
simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@

@@

In the following we will present two example applications for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed. Also for plots we will need packages
\emph{ggplot2} and \emph{gridExtra}.
<<prep>>=
library("model4you")
set.seed(2017)

library("ggplot2")
theme_set(theme_classic())
library("gridExtra")
@

\paragraph{Mathematics exam analysis}
In 2014 first-year business and economics students at the University of Innsbruck
were divided into two examination groups. Group 1 wrote the exam in the morning and
group 2 started after the first group finished.

@@

<<math_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
## compute base model
bmod_math <- lm(pcorrect ~ group, data = MathExam)
lmplot(bmod_math, densest = TRUE)

## extract coefficients and confidence intervals
coefs <- coef(bmod_math)
ci <- confint(bmod_math)
coefs <- cbind(coefs, ci)

## prepare for plotting
digits <- 2
cf <- format(round(coefs, digits), nsmall = digits)
colnams <- colnames(cf)
colnams[1] <- "theta"
gridExtra::grid.table(
  cf, cols = colnams,
  theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
  vp = viewport(x = 0.5, y = 0, width = 0.7, height = 0.4,
                just = c("center", "bottom"))
)
@

<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for math exam.">>=  
## compute tree
(tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2)))
plot(tr_math, terminal_panel = node_pmterminal(tr_math, 
                                               plotfun = lmplot, 
                                               confint = TRUE))
@

<<math_methods>>=  
## predict
newdat <- MathExam[1:5, ]

# terminal nodes
(nodes <- predict(tr_math, type = "node", newdata = newdat))

# response
(pr <- predict(tr_math, type = "pass", newdata = newdat))

# response including confidence intervals, see ?predict.lm
(pr1 <- predict(tr_math, type = "pass", newdata = newdat,
                predict_args = list(interval = "confidence")))
@

<<math_pmodl>>=
forest_math <- pmforest(bmod_math)
pmod_math <- pmodel(forest_math)
head(pmod_math)
@

<<math_dp1, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
dpdat_math <- cbind(pmod_math, MathExam)

ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
  geom_point(alpha = 0.2, size = 1) + 
  geom_smooth(fill = NA, method = "loess") +
  theme(legend.position = "none")

library("ggbeeswarm")
ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
  geom_quasirandom(alpha = 0.5, size = 1) +
  theme(legend.position = "none")
@

<<math_dp2, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
ggplot(dpdat_math, aes(x = attempt, y = group2, color = attempt)) +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none")

ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none")
@

@@

\textcolor{blue}{How the software was implemented, with details of the architecture where relevant. Use of relevant diagrams is appropriate. Please also describe any variants and associated implementation differences.}

@@

\textcolor{blue}{Detail the level of testing that has been carried out on the code (e.g. unit, functional, load etc.), and in which environments. If not already included in the software documentation, provide details of how a user could quickly understand if the software is working (e.g. providing examples of running the software with sample input and output data). }

@@

Should work on all operating systems that run R.
\textcolor{blue}{Please include minimum version compatibility.}

@@

\section*{Additional system requirements}
None.
\textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}

@@

{\bf Archive} \textcolor{blue}{(e.g. institutional repository, general repository) (required ??? please see instructions on journal website for depositing archive copy of software in a suitable repository)}

@@

{\bf Code repository} \textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}

\begin{description}[noitemsep,topsep=0pt]
\item[Name:] R-forge
\item[Persistent identifier:] \url{https://r-forge.r-project.org/projects/partykit/}
\item[Licence:] GPL-2
\item[Date published:] 
\end{description}

@@

\textcolor{blue}{Please describe in as much detail as possible the ways in which the software could be reused by other researchers both within and outside of your field. This should include the use cases for the software, and also details of how the software might be modified or extended (including how contributors should contact you) if appropriate. Also you must include details of what support mechanisms are in place for this software (even if there is no support).}

\section*{Acknowledgements}

\textcolor{blue}{Please add any relevant acknowledgements to anyone else who supported the project in which the software was created, but did not work directly on the software itself.}

@@

\textcolor{blue}{If the software resulted from funded research please give the funder and grant number.}

>>>>

<<setup, echo=FALSE>>=
library("knitr")
knitr::opts_chunk$set(cache = TRUE, message = FALSE, fig.align = "center")

@@

\section*{Abstract}
Typical models estimating treatment effects assume that the treatment effect is
the same for all individuals.  Model-based recursive partitioning allows to
relax this assumption and to estimate stratified treatment effects (model-based
trees) or even personalised treatment effects (model-based forests).  With
model-based trees one can compute treatment effects for different strata of
individuals. The strata are found in a data driven fashion and depend on
characteristics of the individuals.  Model-based random forests allow for a
similarity estimation between individuals in terms of model parameters (e.g.\
intercept and treatment effect). The similarity measure can then be used to
estimate personalised models.  The R package \emph{model4you} implements these
stratified and personalised models with a focus on ease of use and
interpretability so that clinicians and other users can take the model they
usually use for the estimation of the average treatment effect and with a few
lines of code get a visualisation that is easy to understand and interpret.

@@

\paragraph{(1) Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristicum) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data is partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.  
Computing personalised models for all observations in the training data is
simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@

@@

In the following we will present an example application for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed. Also for visualisations we need packages
\emph{ggplot2}, \emph{ggbeeswarm} and \emph{gridExtra}.
<<prep>>=
library("model4you")
set.seed(2017)

library("ggplot2")
theme_set(theme_classic())
library("ggbeeswarm")
library("gridExtra")
@

\paragraph{Mathematics exam analysis:}
In 2014 first-year business and economics students at the University of Innsbruck
were divided into two examination groups. Group 1 wrote the exam in the morning and
group 2 started after the first group finished. The exams for the two groups were 
slightly different. Translated into clinical trial language, the exam groups are
the treatment groups and the two different exams are the treatments.
The data can be accessed and prepared as follows:

@@

To investigate whether the exam was fair, we asses whether the two groups
differ in terms of the percentage of correctly answered exam questions. This
can be done using a simple linear model regressing the percentage points of
correct anwers on the exam group. 
<<math_bmod>>=
bmod_math <- lm(pcorrect ~ group, data = MathExam)
@

The model can be visualised by plotting the estimated densities
(see Figure~\ref{fig:math_bmod_vis}):
<<math_bmod_vis, fig.height = 3, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
lmplot(bmod_math)
@

A tree based on this model can be computed and visualised in only two lines of
code:
<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for math exam.">>=  
tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lmplot))
@

The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \Sexpr{'tests'}. These online tests were
conducted biweekly thrughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well thrughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).

Estimating personalised models is almost as simple as the stratified models:
<<math_pmodl>>=
forest_math <- pmforest(bmod_math)
pmods_math <- pmodel(forest_math)

## model parameters of first 6 students
head(pmods_math)
@

Dependence plots with the group effect (treatment effect) on the y-achsis and the
student characteristics on the x-axis are a good way of visualising the personalised 
models and for getting knowledge about the interactions between student characteristics
and the treatment. Since the percentage of successfull online tests is measured on a
grid, a bee plot possibly shows the relationships even better than the scatter plot
(both shown in Figure~\ref{fig:math_dp1}).
<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
dpdat_math <- cbind(pmods_math, MathExam)

ggplot(dpdat_math, aes(x = tests, y = group2)) +
  geom_point(alpha = 0.2, size = 1) + 
  geom_smooth(fill = NA, method = "loess") +
  theme(legend.position = "none")

ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
  geom_quasirandom(alpha = 0.5, size = 1) +
  theme(legend.position = "none")
@

For the number of previous attempts to pass the exam and the gender box plots, 
bee plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=

ggplot(dpdat_math, aes(x = attempt, y = group2, color = attempt)) +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none")

ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none")
@

With the tools provided by the \emph{model4you} package it is
very simple to create understandable stratified and personalised models and 
compelling visualisations that can be used to communicate theses models.

@@

%\textcolor{blue}{How the software was implemented, with details of the architecture where relevant. Use of relevant diagrams is appropriate. Please also describe any variants and associated implementation differences.}

@@

%\textcolor{blue}{Detail the level of testing that has been carried out on the code (e.g. unit, functional, load etc.), and in which environments. If not already included in the software documentation, provide details of how a user could quickly understand if the software is working (e.g. providing examples of running the software with sample input and output data). }

@@

Should work on all operating systems that run R.
%\textcolor{blue}{Please include minimum version compatibility.}

@@

\section*{Additional system requirements}
None.
%\textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}

@@

{\bf Archive} %\textcolor{blue}{(e.g. institutional repository, general repository) (required ??? please see instructions on journal website for depositing archive copy of software in a suitable repository)}

@@

{\bf Code repository} %\textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}

\begin{description}[noitemsep,topsep=0pt]
\item[Name:] R-forge
\item[Persistent identifier:] \url{https://r-forge.r-project.org/projects/partykit/}
\item[Licence:] GPL-2
\item[Date published:] \textcolor{blue}{dd/mm/yy} %TODO
\end{description}

@@

%\textcolor{blue}{Please describe in as much detail as possible the ways in which the software could be reused by other researchers both within and outside of your field. This should include the use cases for the software, and also details of how the software might be modified or extended (including how contributors should contact you) if appropriate. Also you must include details of what support mechanisms are in place for this software (even if there is no support).}

\section*{Acknowledgements}
%TODO

%\textcolor{blue}{Please add any relevant acknowledgements to anyone else who supported the project in which the software was created, but did not work directly on the software itself.}

@@

%\textcolor{blue}{If the software resulted from funded research please give the funder and grant number.}
